In this paper we extended our previous work \cite{Eades:2017} on a new
foundation of gradual typing to a gradual type system with bounded
quantification called Grady.  We pointed out the existence of explicit
casts in the surface language of gradual type systems that may be used
to extend the expressiveness of the language.  We then proved the
gradual guarantee for Grady.  Finally, we gave several interesting
gradually typed bounded polymorphic programs in Grady's concrete
implementation.

\paragraph{Future work.}  The simplistic nature of Core Grady suggests
that we may extend Grady with even more features.  We plan to study an
extension of Core Grady with dependent types by first extending its
categorical model given in our previous work \cite{Eades:2017}, and
then extending Grady itself.  The idea we have in mind is to make the
static fragment dependent, but the untype fragment remain as it is in
Core Grady.  This is akin to Krishnaswami et al.'s method of
integrating linear and dependent types
\cite{Krishnaswami:2015:ILD:2676726.2676969}.  However, some
interesting problems arise, for example, as we have seen here
non-termination exists in gradually typed programs.  We can type the Y
combinator after all.  Thus, a dependently type gradual type system
would need to employ some method of tracking non-termination, and
preventing non-terminating programs from running in proofs.  This was
exactly the task of the Trelly's project; see \cite{?}.  We believe
that the method developed by Stephanie Weirich and her students
\cite{Casinghino:2014:CPP:2535838.2535883} might be the solution for
gradual typing.  Combing gradual typing and dependent types would
allow for the verification of gradually typed programs.
