Scott \cite{Scott:1980} showed how to model the untyped
$\lambda$-calculus within a cartesian closed category, $\cat{C}$, with
a distinguished object we will call $[[?]]$ -- read as the type of
untyped terms -- such that the object\footnote{We will use the terms
  ``object'' and ``type'' interchangeably.} $[[? ->?]]$ is a retract
of $?$.  That is, there are morphisms $[[squash : (? -> ?)  --> ?]]$
and $[[split : ?  --> (? -> ?)]]$ where $[[squash]];[[split]] = \id :
[[(? -> ?)  --> (? -> ?)]]$\footnote{We denote composition of
  morphisms by $f;g : A \mto C$ given morphisms $f : A \mto B$ and $g
  : B \mto C$.}.  For example, taking these morphisms as terms in the
typed $\lambda$-calculus we can define the prototypical looping term
$(\lambda x.x\,x)(\lambda x.x\,x)$ by $[[(\x:?.H((split x) x))(squash
    (\x:?.H((split x) x)))]]$.

In the same volume as Scott, Lambek \cite{Lambek:1980} showed that
cartesian closed categories also model the typed
$\lambda$-calculus. Suppose we want to model the typed
$\lambda$-calculus with pairs and natural numbers.  That is, given two
types $[[A1]]$ and $[[A2]]$ there is a type $[[A1 x A2]]$, and there
is a type $[[Nat]]$.  Furthermore, we have first and second
projections, and zero and successor functions. This situation can
easily be modeled by a cartesian closed category $\cat{C}$ -- see
Section~\ref{sec:categorical_model} for the details.  Now combine both
of Scott and Lambek's work by adding to $\cat{C}$ the type of untyped
terms $[[?]]$, $[[squash]]$, and $[[split]]$. At this point $\cat{C}$
is a model of both the typed and the untyped $\lambda$-calculus.
However, the two theories are really just sitting side by side in
$\cat{C}$ and cannot really interact much.

Suppose $A$ is an atomic type.  Then we add to $\cat{C}$ the morphisms
$[[box]] : A \mto [[?]]$ and $[[unbox]] :\, ? \mto A$ such that
$[[box]];[[unbox]] = \id : A \mto A$ making $A$ a retract of $[[?]]$.
This is the bridge allowing the typed world to interact with the
untyped one.  We can think of $[[box]]$ as injecting typed data into
the untyped world, and $[[unbox]]$ as taking it back.  Notice that the
only time we can actually get the typed data back out is if it were
injected into the untyped world initially.  In the model this is
enforced through composition, but in the language this will be
enforced at runtime, and hence, requires the language to contain
dynamic typing.  Thus, what we have just built up is a categorical
model that offers a new perspective of how to combine static and
dynamic typing.

Siek and Taha \cite{Siek:2006} define gradual typing to be the
combination of both static and dynamic typing that allows for the
programmer to program in dynamic style without the need for the them
to explicitly insert casts into their programs.  Siek and Taha's
initial paper laid out the first gradually typed $\lambda$-calculus,
but Siek et al. \cite{Siek:2015} layout a refined criteria for what
metatheortic properties a gradual type system should have called the
gradual guarantee.  In this paper we show that our categorical model
leads to a new core casting calculus for gradual type systems.

Siek and Taha's gradually typed $\lambda$-calculus is defined as the
simply typed $\lambda$-calculus with the type of untyped terms $[[?]]$
and the following new application rule:
\begin{mathpar}
\GSiekdruleSXXapp{}
\end{mathpar}
The premise $<<A2 ~ A1>>$ is read, the type $<<A2>>$ is consistent
with the type $<<A1>>$, and is defined in
Figure~\ref{fig:type-consistency}.  If we squint we can see
$[[split]]$, $[[squash]]$, $[[box]]$, and $[[unbox]]$ hiding in the
definition of the previous rules, but they have been suppressed.  We
will show that when one uses either of the two typing rules then one
is really implicitly using a casting morphism built from $[[split]]$,
$[[squash]]$, $[[box]]$, and $[[unbox]]$.  In fact, the consistency
relation $<<A ~ B>>$ can be interpreted as such a morphism.  Then the
typing above can be read semantically as a saying if a casting
morphism exists, then the type really can be converted into the
necessary type.  The premise $<<fun(C) = A1 -> B1>>$ requires that
$<<C>>$ either be $<<A1 -> B1>>$ or $<<C>> = <<?>>$ and
$\mathsf{fun}(C) = [[? -> ?]]$ where $<<A1>> = <<?>>$ and
$<<B1>> = <<?>>$. Again, we can see $[[split]]$ and $[[box]]$ hiding in
the shadows.  When $<<C>> = <<?>>$, then $<<A1>> = <<?>>$ which implies
that $<<A2 ~ A1>>$ will correspond to $[[box]]$, and the premise
$\mathsf{fun}(C) = [[? -> ?]]$ corresponds to using $[[split]]$.  Thus, in this
case the term $<<t1 t2>>$ can be converted into a term in our calculus by
$[[(split (? -> ?) t1) (box A2 t2)]]$.  An interesting point about this
rule is that dynamically typed programs tend toward statically typed programs.
However, the gradual guarantee shows that any program in the gradual type system
can slide either more towards static typing or more towards dynamic typing by
inserting or removing casts.

\paragraph{Contributions.} This paper offers the following contributions:
\begin{itemize}

\item A new categorical model for gradual typing for functional
  languages.  We show how to interpret Siek and Taha's
  \cite{Siek:2015} gradual type system in the categorical model
  outlined above.  

\item We then extract a functional programming language called Simple
  Grady from the categorical model via the Curry-Howard-Lambek
  correspondence.  This is not a gradual type system, but can be seen
  as an alternative core casting calculus in which Siek and Taha's
  gradual type system can be translated to.
  
\item We extend Simple Grady into a new language we call Core Grady that
  has natural numbers, lists, and bounded quantification. In addition,
  natural numbers and lists will contain an eliminator that when
  combined with the Y combinator yields terminating recursion without
  the need for Grady to explicitly contain a fixpoint operator.

\item Finally, we develop a gradually typed surface language called
  Surface Grady and show that it satisfies the gradual guarantee.

\end{itemize}

\paragraph{Related work.} TODO

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "main.tex"  ***
%%% End: ***
