The model we develop here builds on the seminal work of
Lambek~\cite{Lambek:1980} and Scott~\cite{Scott:1980}.
Lambek~\cite{Lambek:1980} showed that the typed $\lambda$-calculus can
be modeled by a cartesian closed category.  In the same volume as
Lambek, Scott essentially showed that the untyped $\lambda$-calculus
is actually typed.  That is, typed theories are more fundamental than
untyped ones.  He accomplished this by adding a single object -- or
type -- $[[?]]$, and two morphisms $[[squash : (? -> ?) -> ?]]$ and
$[[split : ? -> (?  -> ?)]]$, such that, $[[squash]];[[split]] = \id :
[[(? -> ?) -> (?  -> ?)]]$, to a cartesian closed category\footnote{We
  use diagrammatic notation for composition of morphisms.  If $f : [[A
      --> B]]$ and $g : [[B --> C]]$, then their composition is
  denoted by $f;g : [[A --> C]]$.}. At this point he was able to
translate the untyped $\lambda$-calculus into this unityped one.

Categorically, Scott modeled $[[split]]$ and $[[squash]]$ as the
morphisms in a retract within a cartesian closed category -- the same
model as typed $\lambda$-calculus.
\begin{definition}
  \label{def:retract}
  Suppose $\cat{C}$ is a category.  Then an object $A$ is a
  \textbf{retract} of an object $B$ if there are morphisms $i : A \mto
  B$ and $r : B \mto A$ such that $i;r = \id_A$.%% the following diagram commutes:
  %% \[
  %% \bfig
  %% \qtriangle/->`=`->/[A`B`A;i``r]
  %% \efig
  %% \]
\end{definition}
\noindent
Thus, $[[? -> ?]]$ is a retract of $[[?]]$, but we also require that
$[[? x ?]]$ be a retract of $[[?]]$; this is not new, see Lambek and
Scott~\cite{Lambek:1988}.  Putting this together we obtain Scott's
model of the untyped $\lambda$-calculus.
\begin{definition}
  \label{def:model-untyped}
  An \textbf{untyped $\lambda$-model}, $(\cat{C}, ?, \split,
  \squash)$, is a cartesian closed category $\cat{C}$ with a
  distinguished object $?$ and morphisms $\squash : S \mto ?$ and
  $\split : [[?]] \mto S$ making the object $S$ a retract of $?$, where
  $S$ is either $[[? -> ?]]$ or $[[? x ?]]$.
\end{definition}

\begin{theorem}[Scott~\cite{Scott:1980}]
  \label{thm:untyped-lambda-model-sound-complete}
  An untyped $\lambda$-model is a sound and complete model of the untyped $\lambda$-calculus.
\end{theorem}

The untyped $\lambda$ model has the advantage of simplicity, but does not
capture the notion of error handling. To do this, we will introduce the
notion of a monad and its corresponding Kleisli category.

\begin{definition}
  A  \textbf{monad} on a category $\cat{C}$ is a triple $(T,\eta,\mu)$ of a functor $T : \cat{C} \mto \cat{C}$ and two natural transformations $\eta_A : A \mto TA$, $\mu_A : T^2A \mto TA$, such that the following diagrams commute:
  \begin{center}
    \begin{math}
      \bfig
      \square|amma|<500,500>[T^3A`T^2A`T^2A`TA;T_{\mu_A}`\mu_{TA}`\mu_A`\mu_A]
      \efig
      \bfig
      \square|amma|<500,500>[TA`T^2A`T^2A`TA;\eta_{TA}`T_{\eta_A}`\mu_A`\mu_A]
      \morphism(0,500)/=/<500,-500>[TA`TA;]
      \efig
    \end{math}
    \end{center}
\end{definition}

\begin{definition}
  Let $(T,\eta,\mu)$ be a monad on a category $\cat{C}$. The \textbf{Kleisli category}
  of the monad T, denoted $\cat{C}_T$, is the category constructed as follows:
  \begin{enumerate}
    \item The obejcts of $\cat{C}_T$ are those of $\cat{C}$.
    \item The morphisms of $\cat{C}_T$ from A to B are those in $\cat{C}$ from A to TB.
    \item The composition in $\cat{C}_T$ is defined as follows: Let us refer to a
      morphism $f: A \mto TB$ in $\cat{C}$ by $\hat{f} : A \mto B$ when treating it as
      a morphism in $\cat{C}_T$. Then, if $\hat{f} : A \mto B$ and $\hat{g} : B \mto C$
      are morphisms in $\cat{C}_T$, their composition in $\cat{C}_T$, $\hat{f} ;_T \hat{G}$
      is defined to be the morphism in $\cat{C}$ from $A$ to $TC$ given by
      \[
      f; \bar{g}
      \]
  \end{enumerate}
\end{defintion}

To do this, we alter our definition of an
untyped $\lambda$ model to require $\cat{C}$ have all binary (thus all
finite) coproducts, rendering it a \textbf{bicartesian category}. The maybe monad $(T, \eta, \mu)$ is then defined as follows:
\begin{center}
  \begin{math}
    \begin{array}{ll}
      T & = - + 1 \\
      \eta_A : A \mto A + 1 & = i_1 \\
      \mu_A : (A + 1) + 1 \mto A + 1 & = [ \id_{A+1}, i_2] \\
    \end{array}
  \end{math}
\end{center}
It is routine to confirm that these are natural transformations in A,
and that they obey the monad laws; a proof is given in the appendix.

We now require $\split$ be a morphism from $?$ to $S + 1$, and we require
$S$ be a retract of $?$ in the Kleisli category of the monad. That is,
letting $\hat{f}$ denote the Kleisli morphism from $A$ to $B$
corresponding to $f : A \mto B + 1$,
we have, in $\cat{C}_T$,
\begin{center}
  \begin{math}
    \widehat{\squash_S ; \eta_?} ; \widehat{\split_S} = \id_S
  \end{math}
\end{center}
corresponding to the following in $\cat{C}$:
\begin{center}
  \begin{math}
    \squash_S ; \eta_? ; T(\split_S) ; \mu_S = \eta_S
  \end{math}
\end{center}

This is equivalent to requiring that $\squash_S ; \split_S = \eta_S$,
as one may ascertain from the fact that the following diagram commutes:

\begin{center}
  \begin{math}
    \bfig
    \scalefactor{1.2}
    \node A(0,500)[A]
    \node TA1(500,0)[TA]
    \node ?(500,500)[?]
    \node T?(1000,500)[T?]
    \node T2A(1500,500)[T^2A]
    \node TA2(1500,0)[TA]
    \arrow[A`?;\bx_A]
    \arrow|b|[A`TA1;\eta_A]
    \arrow|m|[?`TA1;\unbox_A]
    \arrow[?`T?;\eta_?]
    \arrow[T?`T2A;T\unbox_A]
    \arrow[T2A`TA2;\mu_A]
    \arrow/=/[TA1`TA2;]
    \arrow[TA1`T2A;\eta_{TA}]
    \place(350,350)[3]
    \place(700,300)[1]
    \place(1250,200)[2]
    \efig
  \end{math}
\end{center}

Triangle (1) commutes by naturality of $\eta$, triangle (2) commutes
by the monad laws, and triangle (3) is the retract condition.

Besides this, we want to describe the error case. To this end, we
require for $S_1 \neq S_2$ that $\squash_{S_1} ; \split_{S_2} = \triv_{S_1} ; \textsf{i}_2$.
In the following, $\error_{S_1, S_2}$ will denote $\triv_{S_1} ; \textsf{i}_2$.

Putting it all together, we have the following definition:
\begin{definition}
  \label{def:model-untyped-error}
  An \textbf{untyped $\lambda$-model with error handling}, $(\cat{C}, ?, \split, \squash)$, is a cartesian closed category with coproducts $\cat{C}$, a distinguished object $?$ and morphisms $\squash : S \mto ?$ and $\split : [[?]] \mto S + 1$ for S equal to $[[? x ?]]$, $[[? --> ?]]$ or $? + ?$, related by the following equations:
  \begin{center}
    \begin{math}
      \begin{array}{rll}
        \squash_S ; \split_S & = \eta_S & \\
        \squash_{S_1} ; \split_{S_2} & = \error_{S_1, S_2} & \text{ where } S_1 \neq S_2. \\
      \end{array}
    \end{math}
  \end{center}
\end{definition}

So far we know how to model static types (typed $\lambda$-calculus)
and unknown types (the untyped $\lambda$-calculus).  To make the Grady
languages a bit more interesting we add natural numbers, but we will
need a way to model these in a cartesian closed category.

We model the natural numbers with their (non-recursive) eliminator
using what we call a non-recursive natural number object.  This is a
simplification of the traditional natural number object; see Lambek
and Scott~\cite{Lambek:1988}.
\begin{definition}
  \label{def:SNNO}
  Suppose $\cat{C}$ is a cartesian closed category.  A
  \textbf{non-recursive natural number object (NRNO)} is an object
  $[[Nat]]$ of $\cat{C}$ and morphisms $\mathsf{z} : 1 \mto [[Nat]]$
  and $[[succ]] : [[Nat]] \mto [[Nat]]$ of $\cat{C}$, such that, for
  any morphisms $f : [[Y]] \mto X$ and $g : [[Y x Nat]] \mto X$ of
  $\cat{C}$ there is an unique morphism $\Case_{Y,X} \langle f,g \rangle : [[Y x Nat]] \mto X$
  such that the following hold:
  \[\small
  \setlength{\arraycolsep}{0pt}
  \begin{array}{cccccc}    
    \langle \id_Y, \diamond_Y;\mathsf{z} \rangle;\Case_{Y,X} \langle f,g \rangle = f & \quad &
    \langle \id_Y \times \mathsf{succ} \rangle;\Case_{Y,X} \langle f,g \rangle = g\\
  \end{array}
  \]  
  \noindent
       Informally, the two equations essentially assert that we can
       define $\Case_{Y,X}$ as follows:
       \begin{center}
         \begin{math}
           \setlength{\arraycolsep}{3pt}
           \begin{array}{ccccccccccc}
             {\Case_{Y,X} \langle f,g \rangle\,y\,0 = f\,y}
             & \quad & 
             {\Case_{Y,X} \langle f,g \rangle\,y\,[[(succ n)]] =  g\,y\,n}
           \end{array}
         \end{math}
       \end{center}
\end{definition}

At this point we can model both static and unknown types with natural
numbers in a cartesian closed category, but we do not have any way of
moving typed data into the untyped part and vice versa to obtain
dynamic typing.  We would like to do this in a way that incorporates
error handling.

%[[box C : C --> ?]]
%[[unbox C : ? --> C + 1]]

Naturally, we add $\bx_{C} : C \mto ?$ and $\unbox_{C} : ? \mto C + 1$
such that each atomic type, $[[C]]$, is a retract of $[[?]]$ in
the Kleisli category $\cat{C}_T$ of the maybe monad, as we did for
$\split$ and $\squash$ before. that is,
expanding the Kleisli composition, we require
\begin{center}
\begin{math}
\bx{C} ; \eta_? ; T(\unbox{C}) ; \mu_C = \eta_C.
\end{math}
\end{center}
In fact, this is equivalent to requiring $\bx{C} ; \unbox{C} = \eta_C$
as one can read off from the following diagram.

The only time we can cast $[[?]]$ to another type is if it were boxed up
in the first place.  Combining all of these insights we obtain the
complete categorical model.

\begin{definition}
  \label{def:gradual-lambda-model}
  A \textbf{static-dynamic $\lambda$-model}, $(\cat{T}, \cat{C}, ?,
  \T,\split,$\\ $ \squash, \bx, \unbox, \error)$, where $\cat{T}$ is a
  discrete category with at least two objects $[[Nat]]$ and
  $[[Unit]]$, $\cat{C}$ is a bicartesian closed category with an NRNO,
  $(\cat{C},?,\split,$ $\squash)$ is an untyped $\lambda$-model with
  error handling, $\T : \cat{T} \mto \cat{C}$ is an embedding -- a full
  and faithful functor
  that is injective on objects -- and for every object $A$ of
  $\cat{T}$ there are morphisms $\bx_A : TA \mto ?$ and $\unbox_A : [[?]]
  \mto TA + 1$ making $TA$ a retract of $?$ in the Kleisli category, that
  is, for which $\bx_A : TA \mto ? + 1$ and $\unbox_A : ? \mto TA + 1$
  satisfy $\bx_A ; \unbox_A = \eta_A$. In addition, we require for
  $A \neq B$ that $\bx_A ; \unbox_B = \error_{A,B}$.
  
  %% diagrams commute for any  morphisms $f : [[A --> B]]$ and $g : [[B --> C]]$:

  The following identities hold in such a model:
  \begin{center} 
    \begin{math} \footnotesize
      \setlength{\arraycolsep}{2pt}
      \begin{array}{rll}
        f;\error_{B,C} & = & \error_{A,C}, \text{ where } f : A \mto B\\
        \error_{A,B};f & = & \error_{A,C}, \text{ where } f : B \mto C\\
        \langle \error_{A,B}, f\rangle & = & \error_{A,[[B x C]]}, \text{ where } f : A \mto C\\
        \langle f , \error_{A,C}\rangle & = & \error_{A,[[B x C]]}, \text{ where } f : A \mto B\\
        \curry{\error_{[[A x B]],C}} & = & \error_{A,[[B -> C]]}\\
      \end{array}
    \end{math}
  \end{center}
  Their proofs may be found in the appendix.
\end{definition}
\noindent
We call the category $\cat{T}$ the category of atomic types.  We call
an object, $[[A]]$, \textbf{atomic} iff there is some object $[[A']]$
in $\cat{T}$ such that $[[A]] = \T[[A']]$. Note that we do not
consider $[[?]]$ an atomic type.

Triggering dynamic type errors is a fundamental property of the
criteria for gradually typed languages, and thus, the model must
capture this.  The second coprojection $\err_A : [[1 --> (A + 1)]]$ is combined
with the terminal morphism, $[[triv]]_A : [[A --> 1]]$, which is a
unique morphism guaranteed to exist because $\cat{C}$ is cartesian
closed, to define the morphism $\error_{A,B} : [[A --> B]] + 1$ that
signifies that one tried to unbox or split at the wrong type resulting
in a dynamic type error. If we view morphisms as programs, then
the other equations are congruence rules that trigger a dynamic type
error for the whole program when one of its subparts trigger a dynamic
type error.  The following extends the error equations to the functors
$- \times -$, $- \to -$ and $- + -$:
\begin{lemma}[Extended Errors]
  \label{lemma:extended_errors}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,$\\$\squash, \bx, \unbox,\error)$
  is a gradual $\lambda$-model.  Then the following equations hold:
  \[\footnotesize
  \setlength{\arraycolsep}{2pt}
  \begin{array}{rll}
    f \times \error_{B,C} & = & \error_{[[A x B]],[[C x D]]}, \text{ where } f : A \mto C\\
    \error_{A,C} \times f & = & \error_{[[A x B]],[[C x D]]}, \text{ where } f : B \mto D\\
    f \to \error_{B,C}    & = & \error_{[[A -> B]],[[C -> D]]}, \text{ where } f : C \mto A\\
    \error_{C,A} \to f & = & \error_{[[A -> B]],[[C -> D]]}, \text{ where } f : B \mto D\\
  \end{array}
  \]
\end{lemma}
\begin{proof}
  The following define the morphism part of the functors
  $f \times g : [[(A x B) --> (C x D)]]$, $f \to g : [[(A -> B) --> (C -> D)]]$,
  and $f + g : [[(A + B) --> (C + D)]]$.
  \[ \footnotesize
  \setlength{\arraycolsep}{2pt}
  \begin{array}{rll}
    f \times g & = & \langle [[fst]];f , [[snd]];g \rangle, \\
    & & \,\,\,\,\text{ where } f : A \mto C\text{ and }g : B \mto D\\
    \\
    f \to g & = & \curry{(\id_{[[A -> B]]} \times f);\app_{A,B};g}, \\
    & & \,\,\,\,\text{ where } f : C \mto A\text{ and }g : B \mto D\\
    \\
    f + g & = & \lbrace  f; [[inj1]], g; [[inj2]] \rbrace, \\
    & &  \,\,\,\,\text{ where } f : A \mto C\text{ and }g : B \mto D\\
  \end{array}
  \]
  First, note that $[[fst]] : [[(A x B) --> A]]$, $[[snd]] : [[(A x B) -->
      B]]$, and $\app_{A,B} : [[((A -> B) x A) --> B]]$ all exist by the
  definition of a cartesian closed category.
  
  It is now quite obvious that if either $f$ or $g$ is $\error$ in the
  previous two definitions, then by using the equations from the
  definition of a gradual $\lambda$-model
  (Definition~\ref{def:gradual-lambda-model}) the application of
  either of the functors will result in $\error$.
\end{proof}

As the model is defined it is unclear if we can cast any type to
$[[?]]$, and vice versa, but we must be able to do this in order to
model full dynamic typing.  In the remainder of this section we show
that we can build up such casts in terms of the basic features of our
model.  To cast any type $[[A]]$ to $[[?]]$ we will build casting
morphisms that first take the object $[[A]]$ to its skeleton, and then
takes the skeleton to $[[?]]$.
\begin{definition}
  \label{def:casting-mor}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,\squash, \bx,$\\ $\unbox, \error)$
  is a gradual $\lambda$-model.  Then we call any morphism defined
  completely in terms of $\id$, the functors $- \times -$ and $- \to
  -$, $[[split]]$ and $[[squash]]$, and $[[box]]$ and $[[unbox]]$ a
  \textbf{casting morphism}.
\end{definition}

\begin{definition}
  \label{def:skeleton}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,\squash, \bx,$\\ $\unbox, \error)$
  is a gradual $\lambda$-model.  The \textbf{skeleton} of an object
  $[[A]]$ of $\cat{C}$ is an object $[[S]]$ that is constructed by
  replacing each atomic type in $[[A]]$ with $[[?]]$. Given an object
  $[[A]]$ we denote its skeleton by $[[skeleton A]]$.
\end{definition}
One should think of the skeleton of an object as the supporting type
structure of the object, but we do not know what kind of data is
actually in the structure. For example, the skeleton of the object
$[[Nat]]$ is $[[?]]$, and the skeleton of $[[(Nat x Unit) -> Nat ->
    Nat]]$ is $[[(?  x ?) -> ? -> ?]]$.

The next definition defines a means of constructing a casting morphism
that casts a type $[[A]]$ to its skeleton and vice versa.  This
definition is by mutual recursion on the input type.
\begin{definition}
  \label{def:boxing-unboxing}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,\squash, \bx,$\\ $\unbox, \error)$
  is a gradual $\lambda$-model.  Then for any object $[[A]]$ whose
  skeleton is $[[S]]$ we define the morphisms $[[lbox A]] : [[A -->
      S]]$ and $[[lunbox A]] : [[S --> A]]$ by mutual recursion on
  $[[A]]$ as follows:
  \[\small
  \setlength{\arraycolsep}{1pt}
  \begin{array}{ll|ll}
    \begin{array}{lll}
      [[lbox A]] = [[box A]]\\
    \,\,\,\,\text{when } [[A]]\text{ is atomic}\\
    [[lbox ?]] = \id_{[[?]]}\\
    [[lbox (A1 -> A2)]] = [[lunbox A1]] \to [[lbox A2]]\\
    [[lbox (A1 x A2)]] = [[lbox A1]] \times [[lbox A2]]\\        
    \end{array}
    \\[25px]
    \hline
    \\[-6px]
    \begin{array}{lll}
    [[lunbox A]] = [[unbox A]]\\
    \,\,\,\,\text{when } [[A]]\text{ is atomic}\\
    [[lunbox ?]] = \id_{[[?]]}\\
    [[lunbox (A1 -> A2)]] = [[lbox A1]] \to [[lunbox A2]]\\
    [[lunbox (A1 x A2)]] = [[lunbox A1]] \times [[lunbox A2]]\\        
    \end{array}
  \end{array}
  \]
\end{definition}
\noindent
The definition of both $\widehat{[[box]]}$ or $\widehat{[[unbox]]}$
use the functor $- \to - : \catop{C} \times \cat{C} \mto \cat{C}$
which is contravariant in its first argument, and thus, in that
contravariant position we must make a recursive call to the opposite
function, and hence, they must be mutually defined. Every call to
either $\widehat{[[box]]}$ or $\widehat{[[unbox]]}$ in the previous
definition is on a smaller object than the input object.  Thus, their
definitions are well founded.  Furthermore, $\widehat{[[box]]}$ and
$\widehat{[[unbox]]}$ form a retract between $[[A]]$ and $[[S]]$.
\begin{lemma}[Boxing and Unboxing Lifted Retract]
  \label{lemma:lifted_retract}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,$ $\squash, \bx,$ $\unbox,
  \error)$ is a gradual \\ $\lambda$-model.  Then for any object $[[A]]$,
  $[[lbox A]];[[lunbox A]] = \id_A : [[A --> A]]$.  Furthermore, for
  any objects $[[A]]$ and $[[B]]$ such that $[[A]] \neq [[B]]$, \\
  $[[lbox A]];[[lunbox B]] = \error_{A,B}$.
\end{lemma}
\begin{proof}  
  This proof holds by induction on the form $[[A]]$.  Please see
  Appendix~\ref{subsec:proof_of_lifted_retract} for the complete
  proof.
\end{proof}
\noindent
As an example, suppose we wanted to cast the type $[[(Nat x ?) ->
    Nat]]$ to its skeleton $[[(? x ?) -> ?]]$.  Then we can obtain a
casting morphisms that will do this as follows:
\[
\begin{array}{lll}
  [[lbox ((Nat x ?) -> Nat)]] = ([[unbox Nat]] \times \id_{[[?]]}) \to [[box Nat]]\\
\end{array}
\]

We can also cast a morphism $[[A]] \mto^f [[B]]$ to a morphism
$[[lunbox A]];f;$$[[lbox A]] : [[S1]] \mto [[S2]]$
where $[[S1]] = [[skeleton A]]$ and $[[S2]] = [[skeleton B]]$.  Now if
we have a second
$[[lunbox B]];g;[[lbox C]] : [[S2]] \mto [[S3]]$
then their composition reduces to composition at the typed level:
\begin{center}
  \begin{math}
    \bfig
\square|amma|/->`->`->`<-/<500,500>[
  [[S1]]`
  [[A]]`
  [[S3]]`
  [[C]];
  [[lunbox A]]``
  f;g`
  [[lbox C]]
]

\square(500,0)|amma|/->`->`=`<-/<500,500>[
  [[A]]`
  [[B]]`
  [[C]]`
  [[B]];
  f`
  f;g``
  g]

\square(1000,0)|amma|/->`=`=`<-/<500,500>[
  [[B]]`
  [[S2]]`
  [[B]]`
  [[S2]];
  [[lbox B]]```
  [[lunbox B]]
]
\efig
  \end{math}
\end{center} 
The right most diagram commutes because $[[B]]$ is a retract of
$[[S2]]$, and the left unannotated arrow is the composition $[[lunbox A]];f;$ $g;[[lbox C]]$.  This tells us that we have a functor $\S
: \cat{C} \mto \cat{S}$:
\[
\begin{array}{lll}
  \S[[A]] = [[skeleton A]]\\
  \S (f : [[A --> B]]) = [[lunbox A]];f;[[lbox A]]
\end{array}
\]
where $\cat{S}$ is the full subcategory of $\cat{C}$ consisting of the
skeletons and morphisms between them, that is, $\cat{S}$ is a
cartesian closed category with one basic object $[[?]]$ such that
$(\cat{S},[[?]],[[split]],[[squash]])$ is an untyped $\lambda$-model.
The following turns out to be true.
\begin{lemma}[$\S$ is faithful]
  \label{lemma:S_is_faithful}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,\squash,$\\ $\bx, \unbox, \error)$
  is a gradual $\lambda$-model, and
  $(\cat{S},[[?]],[[split]],[[squash]])$ is the category of skeletons.
  Then the functor $\S : \cat{C} \mto \cat{S}$ is faithful.
\end{lemma}
\begin{proof}
  This proof follows from the definition of $\S$ and Lemma \ref{lemma:lifted_retract}.  For the full proof see
  Appendix~\ref{subsec:proof_of_S_is_faithful}.
\end{proof}
\noindent
Thus, we can think of the functor $\S$ as an injection of the typed
world into the untyped one.

Now that we can cast any type into its skeleton we must show that
every skeleton can be cast to $[[?]]$.  We do this similarly to the
above and lift $[[split]]$ and $[[squash]]$ to arbitrary skeletons.
\begin{definition}
  \label{def:lifted-split-squash}
  Suppose $(\cat{S}, [[?]], [[split]],[[squash]])$ is the category of
  skeletons.  Then for any skeleton $[[S]]$ we define the morphisms
  $[[lsquash S]] : [[S --> ?]]$ and $[[lsplit S]] : [[? --> S]]$ by
  mutual recursion on $[[S]]$ as follows:
  \[ \small
    \begin{array}{l}
      \begin{array}{lll}
        [[lsquash ?]] = \id_?\\
        [[lsquash (S1 -> S2)]] = ([[lsplit S1]] \to [[lsquash S2]]);[[squash ? -> ?]]\\
        [[lsquash (S1 x S2)]] = ([[lsquash S1]] \times [[lsquash S2]]);[[squash ? x ?]]\\
      \end{array}
      \vspace{2px}\\[7px]
      \hline 
      \begin{array}{lll}
        \\ [-6px]
        [[lsplit ?]] = \id_?\\
        [[lsplit (S1 -> S2)]] = [[split ? -> ?]];([[lsquash S1]] \to [[lsplit S2]])\\
        [[lsplit (S1 x S2)]] = [[split ? x ?]];([[lsplit S1]] \times [[lsplit S2]])\\      
      \end{array}
    \end{array}  
  \]
\end{definition}
\noindent
As an example we will construct the casting morphism that casts the
skeleton $[[(? x ?) -> ?]]$ to $[[?]]$:
\[ [[lsquash (? x ?) -> ?]] = ([[split ? x ?]] \to \id_?);[[squash ? -> ?]]. \]

Just as we saw above, splitting and squashing forms a retract.
\begin{lemma}[Splitting and Squashing Lifted Retract]
  \label{lemma:lifted_retract_for_?}
  Suppose $(\cat{S}, [[?]], [[split]],[[squash]])$ is the category of
  skeletons.  Then for any skeleton $[[S]]$, $[[lsquash S]];[[lsplit S]] = \id_S : [[S --> S]]$.
  Furthermore, for
  any skeletons $[[S1]]$ and $[[S2]]$ such that $[[S1]] \neq [[S2]]$,
  $[[lsquash S1]];[[lsplit S2]] = \error_{S1,S2}$.
\end{lemma}
\begin{proof}
  The proof is similar to the proof of the boxing and unboxing lifted
  retract (Lemma~\ref{lemma:lifted_retract}).
\end{proof}
\noindent
There is also a faithful functor from $\cat{S}$ to $\cat{U}$ where
$\cat{U}$ is the full subcategory of $\cat{S}$ that consists of the
single object $[[?]]$ and all its morphisms between it:
\[
\begin{array}{lll}
  \U S = ?\\
  \U (f : [[S1 --> S2]]) = [[lsplit S1]];f;[[lsquash S2]]
\end{array}
\]
This finally implies that there is a functor $\C : \cat{C} \mto
\cat{U}$ that injects all of $\cat{C}$ into the object $[[?]]$.
\begin{lemma}[Casting to $[[?]]$]
  \label{lemma:casting_to_?}
  Suppose $(\cat{T}, \cat{C}, ?, \T, \split,\squash,$\\ $\bx, \unbox, \error)$
  is a gradual $\lambda$-model, $(\cat{S},[[?]],[[split]],[[squash]])$
  is the full subcategory of skeletons, and $(\cat{U},[[?]])$ is the
  full subcategory containing only $[[?]]$ and its morphisms.  Then
  there is a faithful functor $\C = \cat{C} \mto^{\S} \cat{S} \mto^{\U} \cat{U}$.
\end{lemma}
\noindent
In a way we can think of $\C : \cat{C} \mto \cat{U}$ as a forgetful
functor.  It forgets the type information.

Getting back the typed information is harder.  There is no nice
functor from $\cat{U}$ to $\cat{C}$, because we need more information.
However, given a type $[[A]]$ we can always obtain a casting morphism
from $[[?]]$ to $[[A]]$ by $[[(lsplit (skeleton A))]];([[lunbox A]]) :
[[? --> A]]$.  Therefore, we have the following result.
\begin{lemma}[Casting Morphisms to $[[?]]$]
  \label{lemma:casting_morphisms}
  Suppose $(\cat{T}, \cat{C}, ?, \T, $ $\split, \squash, \bx, \unbox, \error)$ is
  a gradual $\lambda$-model, and $[[A]]$ is an object of $\cat{C}$.
  Then there exists casting morphisms from $[[A]]$ to $[[?]]$ and vice
  versa that make $[[A]]$ a retract of $[[?]]$.
\end{lemma}
\begin{proof}
  The two morphisms are as follows:
  \[
  \begin{array}{lll}
    [[Box A]] := [[lbox A]];[[lsquash (skeleton A)]] : [[A --> ?]]\\
    [[Unbox A]] := [[lsplit (skeleton A)]];[[lunbox A]] : [[? --> A]]
  \end{array}
  \]
  \noindent
  The fact the these form a retract between $[[A]]$ and $[[?]]$, and
  raise dynamic type errors holds by Lemma~\ref{lemma:lifted_retract}
  and Lemma~\ref{lemma:lifted_retract_for_?}.
\end{proof}
The previous result has a number of implications.  It completely
brings together the static and dynamic fragments of the gradual
$\lambda$-model, and thus, fully relating the combination of dynamic
and static typing to the past work of Lambek and Scott
\cite{Scott:1980,Lambek:1980}.  It will allow for the definition of
casting morphisms between arbitrary objects.  Finally, from a
practical perspective it will simplify our corresponding type systems
derived from this model, because $[[Box S]] = [[lsquash S]]$ and
$[[Unbox S]] = [[lsplit S]]$ when $[[S]]$ is a skeleton, and hence, we
will only need a single retract in the corresponding type systems.

% subsection the_categorical_model (end)

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "main.tex"  ***
%%% End: ***
