%% The problem:
%%   Not many mathematical models of gradual type systems especially categorical models.
%% Model gives us:
%%   - Connects existing work to gradual typing and the combination of dynamic and static typing
%%   - Can lead to simplifications of the language design: Core Grady does not depend on type consistency
%%   - A categorical model will help in the design of more powerful
%%     gradual type systems, for example, linear types, effects, polymorphism, dependent types, etc.

Gradual typing \cite{Siek:2006,Siek:2015} is a way to combine static
and dynamic typing within the same language.  For example, one can
structure their programs so that the safety critical parts are as
statically typed as possible to catch the most errors at compile time,
while rapidly prototyping other parts of their program using dynamic
typing.

The design of a gradual type system consists of two languages: a core
language and a surface language.  Both languages start as a static
type system with a new type called the unknown type, denoted in this
paper by $[[?]]$, whose inhabitants are all untyped programs.  The
core language then has an explicit mechanism for casting types to and
from the unknown type, and an operational semantics.  However, the
surface language only consists of a type checking algorithm that is
designed so that casts are left implicit, and then after type checking
succeeds a surface language program is translated into the core
language by a cast insertion algorithm.

Programming in the surface language requires the ability to implicitly
cast data between consistent types during eliminations. For example,
$[[(\x:?.(succ (succ x))) 3]]$ should type check with type $[[Nat]]$
even though $[[3]]$ has type $[[Nat]]$ and $[[x]]$ has type
$[[?]]$. Now not every cast should work, for example, $[[(\x:Bool.t)
    3]]$ should not type check, because it is inconsistent to allow
different atomic types to be cast between each other.  Therefore, the
surface language must be able to decide which casts are consistent and
which are not.  This is done by extending the type checking algorithm
with a binary relation called \emph{type consistency} which determines
which types are castable between each other, e.g. every type will be
consistent with the unknown type.

Gradual type systems must satisfy the metatheoretic property
called the \emph{gradual guarantee}.  The gradual guarantee states
that any well-typed program can slide between being more statically
typed or being more dynamically typed by inserting or removing casts
without changing the meaning or the behavior of the program. The
formal statement of the gradual guarantee is given in
Section~\ref{sec:the_gradual_guarantee}.  This property was first
proposed by Siek et al.~\cite{Siek:2015} to set apart systems that
simply combine dynamic and static typing and gradual type systems.

Just as Siek et al. report \cite{Siek:2006,Siek:2007,Siek:2015} there
are a number of programming languages that combine static and dynamic
typing with implicit casting to and from the unknown type, for
example, Boo \cite{Oliveira:2005}, Bigloo
\cite{Bres:2004,Serrano:2002}, Cecil \cite{Chambers:2004}, Visual
Basic .Net, C\# \cite{Meijer:2004}, Professor J
\cite{Gray:2005:FIT:1103845.1094830}, and many more.  In addition,
there are languages that combine static and dynamic typing with
explicit casting only. Abadi et al.~\cite{Abadi:1989} combine dynamic
and static typing by adding a new type called $\mathsf{Dynamic}$ along
with a new case construct for pattern matching on types, and
Henglein~\cite{Henglein:1994} defines the dynamic $\lambda$-calculus
by adding a new type $\mathsf{Dyn}$ to the simply typed
$\lambda$-calculus and then adding primitive casting operations called
tagging and check-and-untag. Please see the introduction to Siek et
al.~\cite{Siek:2015} for a more complete list.

As we can see the combination of dynamic and static typing, as well as
gradual type systems, are important to both industry and academia.
Thus, expanding gradual type systems with new features is an
increasingly important line of research
\cite{Siek:2006,Siek:2007,Garcia:2016,
  Lehmann:2017:GRT:3009837.3009856,
  Jafery:2017:SUR:3093333.3009865}\footnote{There are even more
  examples in the list of accepted papers for ICFP 2017, for example,
  extending gradual type systems with session types and
  polymorphism.}.  Therefore, it is of the utmost importance that we
be able to formally reason about gradual type systems so as to insure
new designs and extensions are correct.  However, there are no
categorical models, in fact not many mathematical models at all, to
aid in the design of new gradual type systems or their extensions.

In this paper we propose a new categorical model of the core casting
calculus of gradual type systems based on the seminal work of
Scott~\cite{Sewell:2010} and Lambek~\cite{Lambek:1988}, thus merging
the categorical model of the statically typed $\lambda$-calculus with
the categorical model of the untyped $\lambda$-calculus into a new
model that combines the two.  Our categorical model leads to a new and
simple type theory that combines dynamic and static typing with
explicit casts called Core Grady.  Furthermore, it is has a less
complex language design compared to existing core languages for
gradual typing, for example, Siek and Taha's system \cite{Siek:2007},
because it does not depend on type consistency where theirs does.  Our
new model also gives a rigorous framework that can be used while
developing new gradual type systems or extensions of existing gradual
type systems.

One strength and main motivation for giving a categorical model to a
programming language is that it can expose the fundamental structure
of the language.  This arises because a lot of the language features
that often cloud the picture go away, for example, syntactic notions
like variables disappear.  This can often simplify things and expose
the underlying structure.  For example, when giving the simply typed
$\lambda$-calculus a categorical model we see that it is a cartesian
closed category, but we also know that intuitionistic logic has the
same model due to Lambek~\cite{Lambek:1980}; on the syntactic side
these two theories are equivalent as well due to
Howard~\cite{Howard:1980}; this is known as the Curry-Howard-Lambek
correspondence.

The previous point highlights one of the most powerful features of
category theory: its ability to relate seemingly unrelated theories.
It is quite surprising that the typed $\lambda$-calculus and
intuitionistic logic share the same model.  Thus, defining a
categorical model for a particular programming language may reveal new
and interesting relationships with existing work.  In fact, one of the
contributions of this paper is the new connection between Scott and
Lambek's work to the new area of gradual typing and combing static and
dynamic typing.

However, that motivation places defining a categorical model as an
after thought.  The programming languages developed here were designed
from the other way around.  We started with the question, how do we
combine static and dynamic typing categorically? Then after developing
the model we use it to push us toward the correct language design.
Reynolds \cite{Brookes:2013} was a big advocate for the use of
category theory in programming language research for this reason.  We
agree with Brookes et al. (from p. 3 of \cite{Brookes:2013}) that the
following quote, originally from
\cite{Reynolds:1995:UFC:199448.199452}, makes this point nicely:
\begin{center}
  \parbox{170pt}{
    \emph{Programming language semanticists should be the obstetricians
    of programming languages, not their coroners.\\
    {\rightline{--- John C. Reynolds}}}}
\end{center}

Categorical semantics provides a powerful tool to study language
extensions.  For example, purely functional programming in Haskell
would not be where it is without the seminal work of Moggi and Wadler
\cite{Moggi:1989,Wadler:1995} on using monads -- a purely categorical
notion -- to add side effects to pure functional programming
languages.  Thus, we believe that developing these types of models for
new language designs and features can be hugely beneficial.

\subsection{Overview}
\label{subsec:main_ideas}
We now give a brief overview of our main results, but from a typed
$\lambda$-calculus perspective, but we will transition to category
theory in Section~\ref{subsec:the_categorical_model}.  Suppose we add
the unknown type, $[[?]]$, and two functions $[[squash]] : [[(? -> ?)
    -> ?]]$ and $[[split]] : [[? -> (? -> ?)]]$ to the simply typed
$\lambda$-calculus with the natural numbers.  Furthermore, we require
that for any program, $[[t]]$, of type $[[? -> ?]]$, we have $[[split
    (squash t)]] \rightsquigarrow t$.  Categorically $[[split]]$ and
$[[squash]]$ define what is called a retract.  Scott \cite{Scott:1980}
showed that this is enough to encode the untyped $\lambda$-calculus
into a statically typed setting:
\begin{center}
  \begin{math} \small
    \begin{array}{lll}
      \setlength{\arraycolsep}{2px}
      \begin{array}{rll}
      [[|x|]]     & = & [[x]]\\
      [[|\x.t|]]  & = & [[\x:?.|t|]]\\      
    \end{array}
      &
      \setlength{\arraycolsep}{2px}
    \begin{array}{rll}
      [[|x t2|]] & = & [[(split x) |t2|]]\\
      [[|t1 t2|]] & = & [[|t1| (squash |t2|)]]
    \end{array}
    \end{array}
  \end{math}
\end{center}
For example, if $\Omega = \lambda x.x\,x$, then $|\Omega| = [[\x:?.H((split x) x)]]$, and
$|\Omega\,\Omega| = |\Omega|\,(\mathsf{squash}\,|\Omega|)$ is the typical diverging term.

We have at this point a typed functional programming language with two
fragments: the statically typed $\lambda$-calculus and the untyped
$\lambda$-calculus.  However, they are just sitting side-by-side.  Now
suppose for any atomic type $[[A]]$, excluding the unknown type, we
add two new functions $[[box A]] : [[A -> ?]]$ and $[[unbox A]] : [[?
    -> A]]$ such that for any term, $[[t]]$, of atomic type $[[A]]$,
we have that $[[unbox A (box A t)]] \rightsquigarrow [[t]]$ -- a
second retract.  This defines the bridge between the typed fragment
and the untyped fragment.  We will show in the next section that both
$[[box]]$ and $[[unbox]]$ can be generalized to arbitrary types, and
thus, they will subsume $[[split]]$ and $[[squash]]$ as well, hence,
reducing all explicit casts to just two functions simplifying the
language even further -- in the rest of this section we only use
$[[box]]$ and $[[unbox]]$.

At this point we have basically built up Core Grady
(Section~\ref{sec:core_grady}) the corresponding type theory to our
categorical model (Section~\ref{subsec:the_categorical_model}).  We
can move statically typed data in between the two fragments. An example
may help solidify the previous point.

Core Grady does not have a primitive notion of recursion, but it is
well-known that we can define the Y combinator in the untyped
$\lambda$-calculus, and hence, in Core Grady. First, we have a full
implementation of every language in this paper
available\footnote{\label{fn:impl} Please see
  \url{https://ct-gradual-typing.github.io/Grady/} for access to the
  implementation as well as full documentation on how to install and
  use it.}. All examples in this paper can be typed and ran in the
implementation, and thus, we make use of Core Grady's concrete syntax
which is very similar to Haskell's and does not venture too far from
the mathematical syntax we will introduce in
Section~\ref{sec:core_grady}.

The definition of the Y combinator in Core Grady is as follows:
\begin{lstlisting}[language=Haskell]
  omega : (? -> ?) -> ?
  omega = \(x : ? -> ?) -> (x (box (? -> ?) x));
  
  fix : (? -> ?) -> ?
  fix = \(f : ? -> ?) ->
            omega (\(x:?) -> f ((unbox (? -> ?) x) x));
\end{lstlisting}
Using $\lstinline{fix}$ we can define the usual arithmetic operations
in Core Grady, but we use a typed version of $\lstinline{fix}$ -- we
have developed many more examples in Core Grady please see the
implementation (Footnote~\ref{fn:impl}).
\begin{lstlisting}[language=Haskell]
  fixNat : ((Nat -> Nat) -> (Nat -> Nat)) -> (Nat -> Nat)
  fixNat = \(f:(Nat -> Nat) -> (Nat -> Nat)) ->
     unbox (Nat -> Nat) (fix (\(y:?)->box (Nat -> Nat) 
                               (f (unbox (Nat -> Nat) y))));

  add : Nat -> Nat -> Nat
  add = \(m:Nat) -> fixNat
       (\(r:Nat -> Nat) ->
        \(n:Nat) -> case n of 0 -> m, (succ n') -> succ (r n'));
        
  mult : Nat -> Nat -> Nat
  mult = \(m:Nat) -> fixNat
       (\(r:Nat -> Nat) ->
        \(n:Nat) -> case n of 0 -> 0, (succ n') -> add m (r n'));
\end{lstlisting}
The function $\lstinline{fixNat}$ is defined so that it does recursion
on the type $[[Nat -> Nat]]$, thus, it must take in an argument,
$f : [[(Nat -> Nat) -> (Nat -> Nat)]]$, and
produce a function of type $[[Nat -> Nat]]$.  However, we
already have $\lstinline{fix}$ defined in the untyped fragment, and
so, we can define $\lstinline{fixNat}$ using $\lstinline{fix}$ by
boxing up the typed data.  This means we must cast $[[f]] : [[(Nat -> Nat) -> (Nat -> Nat)]]$ into a function of type
$[[(? -> ?) -> ?]]$ and we do this by $\eta$-expanding
$[[f]]$ and casting the input and output using $[[box]]$ and
$[[unbox]]$ to arrive at the function
$[[\y:?.H(box (Nat -> Nat) (f (unbox (Nat -> Nat) y)))]] : [[? -> ?]]$.  Finally, we can apply $\lstinline{fix}$, and then unbox its output to the type $[[Nat -> Nat]]$.
Thus, the definition of $\lstinline{fixNat}$ moves typed data to the
untyped fragment using $[[box]]$ and then moves it back using
$[[unbox]]$.

The terms $[[box]]$ and $[[unbox]]$ correspond to explicit casts.  In
Section~\ref{sec:surface_grady} we develop a gradually typed surface
language called Surface Grady that allows for the casts to be left
implicit.  Then we show that both Surface and Core Grady can be
soundly interpreted into our categorical model in
Section~\ref{sec:interpreting_surface_grady_in_the_model}.
Furthermore, we prove the gradual guarantee for Surface and Core Grady
in Section~\ref{sec:the_gradual_guarantee}.

To insure that our categorical model can be used to study other
gradual type systems we show that Siek and Taha's gradually typed
$\lambda$-calculus \cite{Siek:2006,Siek:2015} can also be soundly
modeled by our semantics in
Section~\ref{sec:modeling_siek_and_taha's_gradual_lambda-calculus}.

Finally, while a gradual type system allows for type casts to be left
implicit we show, in
Section~\ref{sec:explicit_casts_in_gradual_type_systems}, that
explicit casts can be derived in the gradually typed surface language,
and using the explicit casts we show that more programs can be typed.
For example, we define a typed fixpoint operator that can only be
defined due to the explicit casts in the gradually typed surface
language.
% subsection main_ideas (end)

%%% Local Variables: ***
%%% mode:latex ***
%%% TeX-master: "main.tex"  ***
%%% End: ***
